A new approach to bias correction in RNA-Seq

### 2011-11-27 response2/response.tex

\documentclass{article}
\usepackage{amsmath}
\usepackage{dcj}

\begin{document}

\section*{Summary of Changes}

\begin{itemize}
\item In response concerns raised by reviewers 2 and 3, we we have reworded the
title and abstract to make clear that our results are most directly applicable
to RNA-Seq, and we are not yet able to make strong assertions regarding
ChIP-Seq.

\item We have taken the first reviewer's suggestion and have highlighted the
strong theoretical results our method provides in the abstract.

\item In response to the first reviewers primary concern, we have clarified the
first paragraph of Section 2.1.1 to make clear that our method does ignore
duplicate reads when training.

\item We have removed our use of the term ``classification'' in the fifth
paragraph of Section 2.1, as it seemed to cause some confusion.

\item We have clarified protocol differences between the Wetterbom and Katze
data sets in Table 1, in response to the second reviewer's question.

\item We have included new data in Supplementary Section 2 showing that the
results presented are not typically sensitive to the values of $p_{\text{max}}$
and $d_{\text{max}}$

\item The description of the procedure we used to compute KL divergence in the
3rd paragraph of Section 3.1 has been rewritten to be more clear.

\item We have corrected an inconsistency in the scale of figures 1 and 4.

\item We have clarified that the values in Table 3 are Pearson correlation
coefficient $r$, and not $r^2$.

\item We have repeated the RT-PCR analysis of 3.3 using version 60 of the
Ensembl gene annotations, rather than version 62, to remain consistent with the
Poisson regression and KL divergence analysis.

\item We have added the proper citation of what we believe to be first published
derivation of the results of Supplementary Section 10.

\time We have cited and made note of in Section 4 a related paper by Zheng, et
al, brought to our attention by reviewer 2.

\item Table 2 was moved to Supplementary Section 3 to preserve space.

\end{itemize}

@@

\begin{quote}
Page 4, line 31, left column. How are the default values of pmax and dmax
determined? Could you provide a table with model complexity, model performance
and running time with different pmax and dmax choices?
\end{quote}

As suggested by Figure 3 in the paper, most edges added to the model are between
adjacent nucleotides, and the in-degree of most positions is small.  We have now
included additional data (Figures 4 and 5 in the Supplement) showing that the
performance of the model on the Mortazavi data set is not greatly effected by
these parameters as long as they are both non-zero. It should not bed assumed
that this necessarily the case with every data set, and so these parameters
default to somewhat arbitrary larger numbers that balance training time with
flexibility.

\begin{quote}
Page 5, line28, right column. This sentence is confusing. Are the duplicated
reads counted or not? Is ``upper half'' referring to the position in the exon or
reads with number of duplicates in the upper 50 percentile?
\end{quote}

We have rewritten the 3rd paragraph in Section 3.1 to provide a more lucid
description of the procedureve.

### 2012-01-03 response2/response.tex

\documentclass{article}
\usepackage{amsmath}
%\usepackage{dcj}

\begin{document}

TODO: brief introduction

We have included below a summary of the changes made in this revision in
addition to specific responses to the points raised by the reviewers.

@@

\section*{Summary of Changes}

\begin{itemize}
\item In response concerns raised by reviewers 2 and 3, we we have reworded the
title and abstract to make clear that our results are most directly applicable
to RNA-Seq, and we are not yet able to make strong assertions regarding
ChIP-Seq.

\item We have taken the first reviewer's suggestion and have highlighted the
strong theoretical results our method provides in the abstract.

\item The first paragraph of Section 2.1.1 has been reworded to make clear that
our method does ignore duplicate reads when training.

@@

\item We have corrected an inconsistency in the scale of figures 1 and 4.

\item We have clarified that the values in Table 3 are Pearson correlation
coefficient $r$, and not $r^2$.

\item We have repeated the RT-PCR analysis of 3.3 using version 60 of the
Ensembl gene annotations, rather than version 62, to remain consistent with the
Poisson regression and KL divergence analysis.

\item In the second paragraph of Section 4, we have added the proper citation of
what we believe to be first published derivation of the results of Supplementary
Section 10.

%\time We have cited and made note of in Section 4 a related paper by Zheng, et
%al, brought to our attention by reviewer 2.

@@

We appreciate this suggestion, and made note of the theoretical results in the
revised abstract of the paper.

\begin{quote}
My only concern is with the sampling described on pg3, column 2, lines 32-47.
In RNA-seq data the distribution of read count can span 5-6 orders of
magnitudes, the top expressed gene accounting for up to 10\% of the data
(especially tissue specific). If you sample at random you are selecting more
frequently from the top, because of this heavy skew in the distribution. You
discuss this issue when measuring KL-divergence in section 3.1, but it does not
appear to affect the training. My main worry is that this will bias the
adjustment to mostly highly expressed genes. How do you take care of this?

One approach would be to simply discard duplicates. In that case how do you
count the sequences, i.e. if two reads map to the same position with one with
sequencing errors, one without do you count them as two sequences or do you
treat them as one and use the sequence of the genome?
\end{quote}

In fact, we do ignore duplicate reads when training the method. This was stated
in the first paragraph of Section 2.2.1 (``we take sequences surrounding
(extending 20 nt to either side, by default) the start positions of a randomly
sampled set of $n/2$ aligned reads, ignoring duplicate reads''), but the reviewer is
correct that this is an important detail, as the method might otherwise be
overfit to a few highly expressed genes. In this revision we have expanded
this paragraph to clarify this decision.

@@

\begin{quote}
pg 3, column 1, lines 42-50 \\
Although BN's are mostly used for classification purposes they are just a
compact representation of a discrete probability distribution, the discussion
of classification is somewhat confusing.
\end{quote}

It was our intention to point out that the training algorithm used is typical of
classification applications, but this point was perhaps needlessly confusing.
We have rewritten the paragraph in question (the fifth paragraph in Section 2.1,
beginning with ``Towards that end...'') to avoid this.

\begin{quote}
supplement pg 7, figure 11. \\
Since you have the 4 replicates from the Trapnell et al. data and you show the
nucleotide frequencies for each replicate, why not run your method independently
on each replicate and compare the modified results.
\end{quote}

We agree that a looking for an increase in correlation of gene abundance
estimates between replicates after bias correction would be an interesting
analysis. Unfortunately, the four RNA-Seq data sets from Trapnell, et al, were
taken from four different biological states and compared without technical or
biological replicates, so a comparison of gene abundance estimates would be
conflated with genuine changes in gene expression.  While we wish to draw
attention to the potential of batch effects in RNA-Seq, a larger survey of
variability of technical bias is outside the intended scope of the paper.

@@

The reviewer is correct that we have by no means established that applying our
method to ChIP-Seq leads to more accurate results, and have instead concentrated
on making this case for RNA-Seq. We have included some very preliminary ChIP-Seq
results, only to suggest that investigating the effects of sequence bias in
ChIP-Seq is a promising future research direction. To avoid the impression that
we are making an overly-broad claim, we have reworded the title and abstract to
reflect that our results relate most readily to RNA-Seq.

\begin{quote}
Moreover, even for RNA-Seq data, only considering biases in base-level read
counts does not seem to be enough for accurate quantitation of transcript level
as shown in (Zheng et al. 2011 BMC Bioinformatics). Features other than the 20
upstream and downstream nucleotides surrounding read starts may also cause
biases in gene/transcript quantitation. The authors may discuss a little more on
their assumptions.
\end{quote}

The Zheng paper makes the assertion that ``the bias patterns across genes were
not adequately addressed by the base-level bias correction methods, and GAM
[Generalized Additive Model] correction is necessary to further remove these
biases.'' This follows from their observation that Hansen's heptamer reweighting
method did not improve correlation bewteen RNA-Seq and RT-PCR. Indeed, our
results are consistent, showing only a slight improvment.

However, the authors apparently did not evaluate this correlation after applying
a more sophisticated model such as the GLM or MART models as propsed by Li. As
we demonstrate in Section 3.3, applying a more flexible model results in a large
increase in correlation, suggesting such an approach should not be hastily
dismissed.

@@

\begin{quote}
Page 2, Figure 1. Wetterbom data does not seem to have much bias in surrounding
nucleotide frequencies. What is the difference between this and Katze data set
that makes such difference? Why do the different Illumina data sets have similar
bias trends whereas ABI data sets do not?
\end{quote}

While both these data sets were generated using the ABI platform, there are
significant differences in the protocol used. Most significantly, the Wetterbom
data was generated using poly-A primers to enrich for mature, polyadenylated
mRNA. In the Katze data, ribosomal RNA was depleted, but polyadenylated
transcripts were not selected for, in an effort to sequence more of the
non-polyadenylated transcriptome. The Kate data set is unique in this aspect:
all three of the Illumina data sets were generated with mRNA-Seq protocols as
well.  We have revised Table 1 to make note of this fact.

\begin{quote}
Page 4, line 31, left column. How are the default values of pmax and dmax
determined? Could you provide a table with model complexity, model performance
and running time with different pmax and dmax choices?
\end{quote}

As suggested by Figure 3 in the paper, most edges added to the model are between
adjacent nucleotides, and the in-degree of most positions is small.  We have now
included additional data (Figures 4 and 5 in the Supplement) showing that the
performance of the model on the Mortazavi data set is not greatly effected by
these parameters as long as they are both non-zero. It should not be assumed
that this necessarily the case with every data set, and so these parameters
default to somewhat arbitrary larger numbers that balance training time with
model flexibility.

@@

\begin{quote}
Page 5, line28, right column. This sentence is confusing. Are the duplicated
reads counted or not? Is ``upper half'' referring to the position in the exon or
reads with number of duplicates in the upper 50 percentile?
\end{quote}

We have rewritten the 3rd paragraph in Section 3.1 to provide a more lucid
description of the procedure.

@@

We use bias to refer to the fact that the estimator in question (RNA-Seq
based measurement of gene abundance) does not, in expectation, agree with the
value we wish to estimate (the true gene abundance). To measure the bias
directly, we would then need to know the true number of copies of each
transcript we wish to quantify and compare RNA-Seq based estimates before and
after correcting for sequence bias. Because exact quantification of transcript
copy number is not feasible at a large scale with any existing technology, we
instead used the largest publically available RT-PCR data set that we are aware
of and demonstrated a substantial increase in the agreement between RNA-Seq and
RT-PCR after correcting for bias in the former (see Section 3.3).

In the example shown in Figure 2, we provided a qualitative illustration of the
large scale quantitative analysis of Section 3.1 and Section 3.2, demonstrating
a reduction in the symptoms of sequence bias across thousands of exons in five
data sets, and in comparison to three previously proposed methods. In sum, we
believe this represents the most complete and rigorous characterization of
sequence bias in RNA-Seq yet presented and is a compelling convergence of
evidence that the accuracy RNA-Seq is improved by correcting for sequence
bias, and that the method we propose is particularily suited to do so.

\begin{quote}
Second, as the authors point out in introduction, another problem of
these biases is that they cannot be assumed to be ``identical between
replicates". How much of the non-identical bias is removed with the
proposed approach?
\end{quote}

We have seen that sequence bias is not always consistent, even in experiments
using similar protocols (see Figure 1, for example), and believe that batch
effects in RNA-Seq are a serious concern. So far, we have concentrated only on
rigorously demonstrating an absolute reduction in sequence bias, and have not
undertaken any systematic comparison of the consistency of this bias between
replicates. Such comparison would be meaningful, but outside of the intended
scope of the paper.

@@

\begin{quote}
Another issue as important as bias reduction is variance reduction. The
bias-variance trade off is unavoidable. By estimating the bias and trying to
adjust for it, more variance is introduced to the system. Is the amount of bias
reduced worth the increased variance, for downstream analysis?
\end{quote}

We agree that it possible to reduce bias and yet introduce so much variance that
that experiments are actually rendered less reliable. In the paper, we attempted
to capture this tradeoff most directly with Figure 5, which demonstrates that,
while uniformity of read coverage is increased ($R^2 > 0$) in the large majority
of exons after applying our method, inevitably it is decreased ($R^2 < 0$) in
some. This suggests that the (BIC) complexity penalty we use during model
selection works as desired, and as as its ansymptotic properties predict,
controlling variance by preventing very dense models from being trained.

### 2012-01-03 seqbias-paper.tex
Final paper, let's hope.

\usepackage{comment}

@@

\section{Motivation:}
Quantification of sequence abundance in RNA-Seq experiments is often conflated
by protocol-specific sequence bias. The exact sources of the bias are unknown,
but may be influenced by PCR amplification, or differing primer affinities and
mixtures, for example. The result is decreased accuracy in many applications,
such as de novo gene annotation and gene or isoform quantification.

@@

In the last few years, RNA-Seq has emerged as a promising alternative to
microarrays in quantifying RNA abundance. But, as microarray technology has
brought with it technical challenges ranging from developing robust
normalization to accounting for cross-hybridization, RNA-Seq presents a new set
of challenges. A significant example is the effect that transcript length has in
biasing quantification, as first explored by \citet{Oshlack2009}, and
subsequently addressed by \cite{Young2010, Gao2011}.

The underlying nucleotide sequence also influences quantification in a manner
that is complex and protocol-specific. In an ideal experiment, the number of
reads mapping to a particular position in the genome is a function of RNA
abundance and should not be additionally dependent on the sequence at that
position. As noted first by \citet{Dohm2008}, this is not the case.  As
illustration, Figure \ref{fig:freqs} plots this nonuniformity in nucleotide
frequencies on five data sets (see Table \ref{tab:datasets}), each using a
different protocol.

These biases may adversely effect transcript discovery, as low level noise may
be over-reported in some regions, and in others, active transcription may be
under-reported. They render untrustworthy comparisons of relative abundance
between genes or isoforms, and any test of differential expression hangs on the
assumption that these biases are identical between replicates, an undesirable
assumption given that the causes of the bias are not well understood.
Additionally, in many tests of differential expression---similar to the effects
of transcript length---a higher read count will result in higher statistical
confidence. It follows that the sensitivity of such a test will also be biased
by sequence, affecting downstream analysis such as gene ontology enrichment
tests.

@@

We begin with a natural model of an RNA-Seq experiment (and one that is often
assumed, whether implicitly or otherwise). The number of reads $x_i$ aligned to
genomic position $i$ is an unbiased estimate of RNA abundance. Furthermore, we
assume reads may be treated as independent and identically distributed samples.
That is, if $N$ reads are generated, and $m_i$ is the event that a generated
read maps to position $i$, then
$$ E[x_i] = N \Pr[m_i] $$

The experiment may be considered unbiased  with regards to sequence if, having
observed the nucleotide sequence $s_i$ surrounding position $i$, the expected
number of reads sampled from position $i$ is independent of $s_i$, i.e., if

$$ E[ x_i | s_i ] = N \Pr[ m_i | s_i ] = N \Pr[ m_i ] = E[ x_i ] $$
From Bayes' rule,
$$ \Pr[ m_i | s_i ] = \frac{ \Pr[ s_i | m_i ] \Pr[ m_i ] }{ \Pr[ s_i ] } $$
This suggests a natural scheme in which observations may be reweighted to
correct for bias.  First, define the \emph{sequence bias} $b_i$ at position $i$
as $ b_i = \Pr[ s_i ] /  \Pr[ s_i | m_i ]  $.

@@

Towards that end, we propose training a Bayesian network using examples of
foreground and background sequences. By training the model discriminatively and
penalizing model complexity, we can avoid a model that is over-parametrized,
excluding parameters that are insufficiently informitive in discriminating
between foreground and background. The Bayesian network can then be used to
evaluate sequence probability, and thus bias, at any genomic position. Figure
\ref{fig:overview} gives a schematic overview of the proposed model.

\begin{figure}
\centerline{\includegraphics[width=0.37\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the model is evaluated
at each position within a locus, predicting bias. The predicted bias can then be
used to adjust read counts, as in (e). In (d) and (e) we show the results of
this method applied to the $3'$ UTR of Apoa2, using data from
\citet{Mortazavi2008}. In bias coefficients predicted across 10 million positions
of chromosome 1, the log10 bias of 95\% of the positions were between -1.14 and
0.63, suggesting that most adjustments are not large. The $R^2$ measure,
detailed in Section 3.2, gives the relative increase in log-likeliood under a
uniform sampling model, after correcting for bias, with 1.0 indicating a perfect
fit, and the score of 0.38 here indicating a significant increase.
}
\label{fig:overview}
\end{figure}

However, we have so far ignored one complication: the RNA abundance that we wish
to estimate is not itself independent of the nucleotide sequence. Notably,
exonic DNA tends to be more GC-rich than intergenic DNA. If background sequences
are sampled uniformly from the genome we run the risk of incorrectly adjusting
for biological sequence bias, rather than technical sequence bias.  To avoid
this, we propose using paired training data. Each foreground training
sequence is paired with a background sequence taken from a nearby position
that is likely to have similar abundance and general nucleotide composition.
Alternatively, we could pair foreground samples with background samples from
within the same transcript, but we prefer to avoid dependence on existing
gene annotations as such a model would be limited to RNA-Seq analysis in which
trustworthy and reasonably complete gene annotations are available.

@@

\subsubsection{Sampling}

The model is trained on $n$ sequences, one half labeled as foreground, the other
background, sampled from the reference genome. To obtain the foreground
sequences, we take sequences surrounding (extending 20 nt to either side, by
default) the start positions of a randomly sampled set of $n/2$ aligned reads.
To avoid the risk of the method being overfit to reads deriving from a few
highly expressed genes, we ignore duplicate reads, which we define as tow reads
mapping to the same location in the genome. The nucleotide sequence is taken
from the genome, rather then the reads themselves, allowing us to include
positions outside of the read.

@@

\subsubsection{Training}

To determine the structure and parameters of the Bayesian network, we use a
hill-climbing approach similar to the algorithm described by \citet{Grossman2004}.
The network structure is determined by greedily optimizing the conditional
log-likelihood:
\begin{align*}
\ell &= \sum_{i=1}^{n} \log \Pr( x_i | s_i ) 
=
\sum_{i=1}^{n} \log \frac{ \Pr(s_i | x_i) \Pr( x_i ) }{
\sum_{x \in \{0,1\}} \Pr( s_i | x ) \Pr(x) } \\
\end{align*}
where $\Pr(x)$ is flat (i.e.  $\Pr( x = 0 ) = \Pr( x = 1 ) =
0.5$) since we sample foreground and background positions equally.

@@

At each step of the optimization procedure, every possible edge or position
addition, removal, or edge reversal that produces a valid network (i.e., does not
introduce a cycle) is evaluated, and the alteration that increases the score
$\ell'$ the most is kept.  This process is repeated until a local maximum is
found, in which no single alteration to the network will increase the score.
Given the network structure, the parameters are estimated directly from the
observed nucleotide frequencies in the training data.

The run time of the training procedure is further reduced in practice by
imposing the following two restrictions on the structure of the network, First,
the in-degree (i.e., number of parents) of any node must be less than some
number $p_{\text{max}}$. Secondly, for all edges $(i,j)$, $|j - i| \le
d_{\text{max}}$ for some number $d_{\text{max}}$.  This encodes the assumption
that distant nucleotides are effectively independent. We choose $p_{\text{max}}
= 4$ and $d_{\text{max}} = 10$, as reasonable default values.

Along with $p_{\text{max}}$ and $d_{\text{max}}$, the time required to train the
model is controlled by the number of reads used (i.e., $n$) and the number of
positions surrounding the read start to consider. Increasing any of these
parameters results in a  solution that is potentially better, at
the expense of increased training time.

@@

Because we cannot observe directly the underlying RNA-abundance, our evaluation
strategy relies on testing three assumptions we make of an ideal, unbiased experiment.
\begin{enumerate}
\item Positional nucleotide frequencies (as in Figure \ref{fig:freqs}), measured
from reads within exons, should not differ greatly from frequencies measured by
sampling uniformly within the same exons.
\item Read counts across a single exon should follow, approximately, a Poisson
process.
\item Adjusting for bias in RNA-Seq should increase the agreement between RNA-Seq and
an other methods of quantification.
\end{enumerate}

@@

Testing was performed by cross-validation. Each method was trained on data taken
from chromosomes 1--8 of the genome from which the reads were mapped (including
chromosomes 2a and 2b of the Chimpanzee genome). For evaluation, we drew a set
of long, highly expressed exons from the remaining chromosomes. In particular,
for each reference sequence, beginning with the set of exons annotated by
Ensembl release 60 \cite{Hubbard2009}, we removed any exons with known alternate
splice sites, then chose the top 1000 exons by read count, restricting ourselves
to those at least 100 nt long.

@@

This measure can be interpreted as the improvement in fit over the null model,
with $R^2 = 1$ indicating a perfect fit, occurring when the model being
evaluated achieves a likelihood of 1. Smaller number indicate an increasingly
worse fit, with $R^2 = 0$ representing no improvement over the null model, and
$R^2 = 0.5$, for example, indicating the model has a log-likelihood equal to
half that of the null model (a large improvement, corresponding to, for example,
the likelihood increasing over 100-fold if the initial log-likelihood was -9.6,
which is the mean per-position log-likehood under the null model). This measure
can also take on values less than 0, indicating that the model has worse fit
than the null model (i.e., when adjusting read counts according the bias
coefficients leads to less uniform read coverage).

@@

We used sequencing data previously published by \citet{Au2010} to evaluate the
effect bias correction has on correlation to measurements made by TaqMan RT-PCR,
made available by the the Microarray Quality Control project \citep{Shi2006}. The
RNA-Seq data shows a pattern of bias similar to that seen in the other samples
sequenced on an Illumina platform (see Supplementary Section 6). This evaluation
does not rely on an assumption that qRT-PCR necissarily more accurate than
RNA-Seq based quantification, only that qRT-PCR is not biased in the same way as
the RNA-Seq data.

@@

In our results, estimating initial heptamer frequencies was not seen to be as
effective as the other models, even when data generated using random hexamer
priming was used. A possible explaination is, given the large number of
parameters needed to estimate heptamer frequencies ($4^{7} = 16,383$), these
parameters are estimated with less accuracy than in models requiring fewer
parameters. Yet, we trained the 7mer model on a minimum of 1.9 million reads
(see Supplementary Section 3), a number which our theoretical results, detailed
in Supplementary Section 10, suggest should lead to accurate estimates. A perhaps
more significant factor is that this method does not capture bias outside of the
initial heptamer, though many data sets clearly are affected by bias in other
positions.  Thus to improve the performance, it seems necessary to incease the
size of the $k$-mers being considered. Yet, exponentially more reads would be
required for an accurate estimate since the accuracy of the model, as quantified
by its KL divergence from the true distribution, is $(r-1)/2n$, where $r=4^k$ is
the number of parameters that must be estimated (See Supplementary Section 10).

@@

@@

To determine the extent to which PCR amplification is responsible for the
observed bias, we evaluated data from the FRT-Seq method proposed by
\citet{Mamanova2010}.  FRT-Seq avoids the PCR amplification step during library
preparation with reverse transcription occuring on the flowcell surface. We
observed that this data is not free from sequence bias, yet unlike other data
generated on the Illumina platform, it appears to be effected only by relatively
few positions adjacent to the read start (see Supplementary Section 8).

>>>>

@@

\section{Motivation:}
Quantification of sequence abundance in RNA-Seq experiments is often conflated
by protocol-specific sequence bias. The exact sources of the bias are unknown,
but may be influenced by PCR amplification, or differing primer affinities and
mixtures, for example. The result is decreased accuracy in many applications,
such as de novo gene annotation and transcript quantification.

@@

In the last few years, RNA-Seq has emerged as a promising alternative to
microarrays in quantifying RNA abundance. But, as microarray technology has
brought with it technical challenges ranging from developing robust
normalization to accounting for cross-hybridization, RNA-Seq presents a new set
of challenges. As first noted by \citet{Dohm2008}, a particular challenge is the
often complex and protocol-specific influence of nucleotide sequence on
quantification.

In an ideal experiment, the number of RNA-Seq reads mapping to a particular
position in the genome is a function of RNA abundance and should not be
additionally dependent on the sequence at that position. Yet, this is not the
case. As illustration, Figure \ref{fig:freqs} plots this nonuniformity in
nucleotide frequencies on five data sets (see Table \ref{tab:datasets}), each
using a different protocol.

These biases may adversely effect transcript discovery, as low level noise may
be over-reported in some regions, and in others, active transcription may be
under-reported. They render untrustworthy comparisons of relative abundance
between genes or isoforms, and any test of differential expression hangs on the
assumption that these biases are identical between replicates, an undesirable
assumption given that the causes of the bias are not well understood.
Additionally, in many tests of differential expression higher read count will
result in higher statistical confidence. It follows that the sensitivity of such
a test will also be biased by sequence, affecting downstream analysis such as
gene ontology enrichment tests.

@@

We begin with a natural model of an RNA-Seq experiment (and one that is often
assumed, whether implicitly or otherwise). The number of reads $x_i$ aligned to
genomic position $i$ is an unbiased estimate of RNA abundance. Furthermore, we
assume reads may be treated as independent and identically distributed samples.
That is, if $N$ reads are generated, and $m_i$ is the event that a generated
read maps to position $i$, then $ E[x_i] = N \Pr[m_i] $.

The experiment may be considered unbiased with regards to sequence if, having
observed the nucleotide sequence $s_i$ surrounding position $i$, the expected
number of reads sampled from position $i$ is independent of $s_i$, i.e., if
$$ E[ x_i | s_i ] = N \Pr[ m_i | s_i ] = N \Pr[ m_i ] = E[ x_i ] $$
From Bayes' rule,
$$ \Pr[ m_i | s_i ] = \frac{ \Pr[ s_i | m_i ] \Pr[ m_i ] }{ \Pr[ s_i ] } $$
This suggests a natural scheme in which observations may be reweighted to
correct for bias.  First, define the \emph{sequence bias} $b_i$ at position $i$
as $ b_i = \Pr[ s_i ] /  \Pr[ s_i | m_i ]  $.

@@

Towards that end, we propose training a Bayesian network using examples of
foreground and background sequences. By training the model discriminatively and
penalizing model complexity, we can avoid a model that is over-parametrized,
excluding parameters that are insufficiently informative in discriminating
between foreground and background. The Bayesian network can then be used to
evaluate sequence probability, and thus bias, at any genomic position. Figure
\ref{fig:overview} gives a schematic overview of the proposed model.

\begin{figure}
\centerline{\includegraphics[width=0.37\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the model is evaluated
at each position within a locus, predicting bias. The predicted bias can then be
used to adjust read counts, as in (e). In (d) and (e) we show the results of
this method applied to the $3'$ UTR of Apoa2, using data from
\citet{Mortazavi2008}. In bias coefficients predicted across 10 million positions
of chromosome 1, the log10 bias of 95\% of the positions were between -1.14 and
0.63, suggesting that most adjustments are not large. The $R^2$ measure,
detailed in Section 3.2, gives the relative increase in log-likeliood under a
uniform sampling model, after correcting for bias, with 1.0 indicating a perfect
fit, and the score of 0.38 here indicating a significant increase.}
\label{fig:overview}
\end{figure}

We have so far ignored one complication: the RNA abundance that we wish
to estimate is not itself independent of the nucleotide sequence. Notably,
exonic DNA tends to be more GC-rich than intergenic DNA. If background sequences
are sampled uniformly from the genome we run the risk of incorrectly adjusting
for biological sequence bias, rather than technical sequence bias.  To avoid
this, we propose using paired training data. Each foreground training
sequence is paired with a background sequence taken from a nearby position
that is likely to have similar abundance and general nucleotide composition.
Alternatively, we could pair foreground samples with background samples from
within the same transcript, but we prefer to avoid dependence on existing
gene annotations.

@@

\subsubsection{Sampling}
The model is trained on $n$ sequences, one half labeled as foreground, the other
background, sampled from the reference genome. To obtain the foreground
sequences, we take sequences surrounding (extending 20 nt to either side, by
default) the start positions of a randomly sampled set of $n/2$ aligned reads.
To avoid the risk of the method being overfit to reads deriving from a few
highly expressed genes, we ignore duplicate reads, which we define as two reads
mapping to the same location in the genome. The nucleotide sequence is taken
from the genome, rather than the reads themselves, allowing us to include
positions outside of the read.

@@

\subsubsection{Training}
To determine the structure and parameters of the Bayesian network, we use a
hill-climbing approach similar to the algorithm described by \citet{Grossman2004}.
The network structure is determined by greedily optimizing the conditional
log-likelihood:
\begin{align*}
\ell &= \sum_{i=1}^{n} \log \Pr[ x_i | s_i ]
=
\sum_{i=1}^{n} \log \frac{ \Pr[s_i | x_i] \Pr[ x_i ] }{
\sum_{x \in \{0,1\}} \Pr[ s_i | x ] \Pr[x] } \\
\end{align*}
where $\Pr[x]$ is flat (i.e.  $\Pr[ x = 0 ] = \Pr[ x = 1 ] =
0.5$) since we sample foreground and background positions equally.

@@

At each step of the optimization procedure, every possible edge or position
addition, removal, or edge reversal that produces a valid, acyclic network
is evaluated, and the alteration that increases the score
$\ell'$ the most is kept.  This process is repeated until a local maximum is
found, in which no single alteration to the network will increase the score.
Given the network structure, the parameters are estimated directly from the
observed nucleotide frequencies in the training data.

The run time of the training procedure is further reduced in practice by
imposing the following two restrictions on the structure of the network, First,
the in-degree (i.e., number of parents) of any node must be less than some
number $p_{\text{max}}$. Secondly, for all edges $(i,j)$, $|j - i| \le
d_{\text{max}}$ for some number $d_{\text{max}}$. This latter rule encodes the
assumption that distant nucleotides are effectively independent. We choose
$p_{\text{max}} = 4$ and $d_{\text{max}} = 10$, as reasonable default values
(see Supplementary Section 2).

@@

Because we cannot observe directly the underlying RNA-abundance, our evaluation
strategy relies on testing three assumptions we make of an ideal, unbiased
RNA-Seq experiment.
\begin{enumerate}
\item Positional nucleotide frequencies (as in Figure \ref{fig:freqs}), measured
from reads within exons, should not differ greatly from frequencies measured by
sampling uniformly within the same exons.
\item Read counts across a single exon should follow, approximately, a Poisson
process.
\item Adjusting for bias in RNA-Seq should increase the agreement between RNA-Seq and
an other methods of quantification.
\end{enumerate}

@@

Each method was trained on data taken
from chromosomes 1--8 of the genome from which the reads were mapped (including
chromosomes 2a and 2b of the Chimpanzee genome). For evaluation, we drew a set
of long, highly expressed exons from the remaining chromosomes. In particular,
for each reference sequence, beginning with the set of exons annotated by
Ensembl release 60 \cite{Hubbard2009}, we removed any exons with known alternate
splice sites, then chose the top 1000 exons by read count, restricting ourselves
to those at least 100 nt long.

@@

This measure can be interpreted as the improvement in fit over the null model,
with $R^2 = 1$ indicating a perfect fit, occurring when the model being
evaluated achieves a likelihood of 1. Smaller number indicate an increasingly
worse fit, with $R^2 = 0$ representing no improvement over the null model, and
$R^2 = 0.5$, for example, indicating the model has a log-likelihood equal to
half that of the null model (a large improvement, corresponding to, for example,
the likelihood increasing over 100-fold if the initial log-likelihood was -9.6,
which is the mean per-position log-likehood under the null model). This measure
has the added advantage that it can take on values less than 0, indicating that
the model has worse fit than the null model (i.e., when adjusting read counts
according the bias coefficients leads to less uniform read coverage).

@@

We used sequencing data previously published by \citet{Au2010} to evaluate the
effect bias correction has on correlation to measurements made by TaqMan RT-PCR,
made available by the the Microarray Quality Control project \citep{Shi2006}. The
RNA-Seq data shows a pattern of bias similar to that seen in the other samples
sequenced on an Illumina platform (see Supplementary Section 6). This evaluation
does not rely on an assumption that qRT-PCR is necessarily more accurate than
RNA-Seq based quantification, only that qRT-PCR is not biased in the same way as
the RNA-Seq data.

@@

In our results, estimating initial heptamer frequencies was not seen to be as
effective as the other models, even when data generated using random hexamer
priming was used. A possible explanation is, given the large number of
parameters needed to estimate heptamer frequencies ($4^{7} = 16,383$), these
parameters are estimated with less accuracy than in models requiring fewer
parameters. Yet, we trained the 7mer model on a minimum of 1.9 million reads
(see Supplementary Section 3), a number which theoretical results,
following from work by \citet{Birch1964} and included in Supplementary Section
10 for completeness, suggest should lead to accurate estimates

A perhaps more significant factor is that this method does not capture bias
outside of the initial heptamer, though many data sets clearly are affected by
bias in other positions.  Thus to improve the performance, it seems necessary to
increase the size of the $k$-mers being considered. However, exponentially more reads
would be required for an accurate estimate since the accuracy of the model, as
quantified by its KL divergence from the true distribution, is $(r-1)/2n$, where
$r=4^k$ is the number of parameters that must be estimated (See Supplementary
Section 10).

@@

During the review of this manuscript, another method addressing sequence bias in
RNA-Seq was published by \citet{Zheng2011}. Rather than fitting a model of the
specific base-level sequence bias surrounding read start, they propose making
adjustments according to summary statistics at the gene level. Such an approach
is disadvantaged in its inability to model the very specific pattern of sequence
bias we have observed. Yet, such an approach is efficient, and though we have
not yet evaluated it, claimed to be effective.

@@

To determine the extent to which PCR amplification is responsible for the
observed bias, we evaluated data from the FRT-Seq method proposed by
\citet{Mamanova2010}. FRT-Seq avoids the PCR amplification step during library
preparation with reverse transcription occuring on the flowcell surface. We
observed that this data is not free from sequence bias, yet unlike other data
generated on the Illumina platform, it appears to be effected only by relatively
few positions adjacent to the read start (see Supplementary Section 8). Other
protocol improvements might further reduce sequence bias. Notably, promising work
by \citet{Jayaprakash2011} proposes a pooled adapter strategy to deal with this
issue in small RNA sequencing experiments.
